Parameter Recovery Validation

### 2025-06-14 nbs/concepts/parameter-recovery-validation/index.qmd
docs(parameter-recovery-validation): justify introduction of stick-breaking procedure
Signed-off-by: Cameron Smith <cameron.ray.smith@gmail.com>

**Proof:**
1. **Fixed anchor constraint**: The scaled system requires $t^*_1 = \lambda \cdot 0 = 0$ and $t^*_N = \lambda T_{\max}$
2. **Process duration independence**: Gene parameters $t^*_{0,\text{on}i}, \delta^*_i$ are sampled independently of $T_{\max}$ and do not scale
3. **Asymmetric scaling**: Scaling $T_{\max} \rightarrow \lambda T_{\max}$ and $t^*_j \rightarrow \lambda t^*_j$ changes the relative positioning of gene events within the process timeline
4. **Likelihood dependence**: The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$, which change under asymmetric scaling

@@

>>>>

1. **Fixed anchor constraint**: The scaled system requires $t^*_1 = \lambda \cdot 0 = 0$ and $t^*_N = \lambda T_{\max}$
2. **Process duration independence**: Gene parameters $t^*_{0,\text{on}i}, \delta^*_i$ are sampled independently of $T_{\max}$ and do not scale
3. **Asymmetric scaling**: Scaling $T_{\max} \rightarrow \lambda T_{\max}$ and $t^*_j \rightarrow \lambda t^*_j$ changes the relative positioning of gene events within the process timeline
4. **Likelihood dependence**: The piecewise analytical solutions depend on comparisons like $t^*_j$ vs. $t^*_{0,\text{on}i}$, which change under asymmetric scaling

@@

#### **Mathematical and Biological Foundations of Stick-Breaking for Temporal Coordinates**

The stick-breaking construction for temporal coordinates addresses fundamental challenges in modeling cellular developmental progression and provides the mathematically natural solution to the constrained latent temporal assignment problem.

**1. The Fundamental Problem Being Solved**

**Physical vs. Biological Time Mismatch:**

Single-cell RNA sequencing captures all cells simultaneously at a fixed experimental time $t_{\text{experiment}}$, yet each cell exists at a different stage of the underlying biological process. This creates a fundamental mismatch:

- **Physical observation**: All cells measured at $t_{\text{experiment}}$
- **Biological reality**: Each cell $j$ at developmental stage $t^*_j$ within process duration $[0, T_{\max}]$

**Constrained Latent Temporal Assignment Problem:**

The inference challenge becomes assigning temporal coordinates $\{t^*_j\}_{j=1}^N$ that satisfy multiple biological constraints:

\begin{align}
\mathcal{C}_{\text{boundary}} &: t^*_1 = 0, \quad t^*_N = T_{\max} \quad \text{(process boundaries)} \\
\mathcal{C}_{\text{ordering}} &: t^*_1 \leq t^*_2 \leq \cdots \leq t^*_N \quad \text{(developmental progression)} \\
\mathcal{C}_{\text{continuity}} &: \|x_i - x_j\| \text{ small} \Rightarrow |t^*_i - t^*_j| \text{ small} \quad \text{(state-time proximity)}
\end{align}

where $x_i, x_j$ represent gene expression states and $\mathcal{C}_{\text{continuity}}$ enforces the differential equation requirement that cells close in expression space should be close in developmental time.

**2. Mathematical Object Definition**

**Constrained Order Statistics Space:**

Stick-breaking provides a probability distribution over the constrained order statistics space:

$$\mathcal{T}_N^{T_{\max}} = \{(t^*_1, t^*_2, \ldots, t^*_N) \in \mathbb{R}^N : 0 = t^*_1 \leq t^*_2 \leq \cdots \leq t^*_N = T_{\max}\}$$

This space is isomorphic to the $(N-2)$-simplex with fixed boundary conditions, enabling efficient sampling and probability computation.

**Probability Density with Jacobian Transformation:**

The stick-breaking construction transforms independent Beta random variables to ordered temporal coordinates:

\begin{align}
\xi_j &\sim \text{Beta}(1, N-j) \quad \text{for } j \in \{2, \ldots, N-1\} \\
t^*_j &= t^*_{j-1} + \xi_{j-1}(T_{\max} - t^*_{j-1}) \quad \text{(recursive construction)}
\end{align}

The joint density on $\mathcal{T}_N^{T_{\max}}$ becomes:

$$p(t^*_2, \ldots, t^*_{N-1} | T_{\max}) = \frac{(N-2)!}{T_{\max}^{N-2}} \prod_{j=2}^{N-1} (N-j) \quad \text{for } (t^*_1, \ldots, t^*_N) \in \mathcal{T}_N^{T_{\max}}$$

This uniform density over the constrained space reflects maximum entropy subject to biological constraints.

**3. State Space Proximity Preservation**

**Ordering Constraint and Biological Continuity:**

The ordering constraint $t^*_1 \leq t^*_2 \leq \cdots \leq t^*_N$ directly enforces the biological assumption that cells close in expression state should be close in developmental time. This constraint ensures that the temporal manifold respects the underlying differential equation dynamics.

**Variable Spacing Property:**

The stick-breaking construction exhibits adaptive spacing that accommodates heterogeneous cell density:

$$\mathbb{E}[t^*_{j+1} - t^*_j | T_{\max}] = \frac{T_{\max}}{N+1} \cdot \frac{1}{N-j}$$

This variable spacing allows:
- **Dense regions**: Smaller temporal intervals where many cells cluster in expression space
- **Sparse regions**: Larger temporal intervals where few cells exist
- **Adaptive resolution**: Automatic adjustment to local cell density

**State Space Distance Integration:**

The Beta parameters can be adapted to incorporate state space distance information:

$$\xi_j \sim \text{Beta}(\alpha_j, \beta_j) \quad \text{where } \alpha_j, \beta_j \propto f(\|x_{j+1} - x_j\|)$$

enabling data-driven temporal spacing that respects expression similarity.

**4. Information-Theoretic Optimality**

**Maximum Entropy Principle:**

Stick-breaking with uniform Beta$(1, N-j)$ parameters is the **maximum entropy distribution** subject to:
- Fixed endpoints: $t^*_1 = 0$, $t^*_N = T_{\max}$
- Ordering constraints: $t^*_j \leq t^*_{j+1}$
- Exchangeability: Symmetric treatment of intermediate cells

**Minimal Assumptions:**

This maximum entropy property ensures **minimal assumptions** beyond biological requirements:
- No artificial temporal clustering
- No preferred developmental stages
- Maximal uncertainty preservation where data provides no constraints

**Information Preservation:**

The construction preserves **maximal information** about temporal relationships by avoiding arbitrary distributional assumptions that could bias inference toward specific temporal patterns.

**5. Why Alternative Approaches Fail**

**Random Temporal Assignment:**

Random assignment of temporal coordinates $t^*_j \sim \text{Uniform}(0, T_{\max})$ violates differential equation continuity assumptions:
- Cells with similar expression states could be assigned vastly different developmental times
- Discontinuous jumps in the temporal manifold create unrealistic dynamics
- Loss of biological interpretability for developmental progression

**Truncated/Constrained Distributions:**

Direct sampling from truncated distributions over ordered statistics creates computational problems:
- Complex normalization constants that scale exponentially with $N$
- Inefficient rejection sampling with low acceptance rates
- Numerical instability for large cell populations

**Uniform Distributions:**

Uniform spacing $t^*_j = j \cdot T_{\max}/N$ truncates the dynamical range needed for meaningful exponential convergence:
- Fixed spacing cannot adapt to heterogeneous cell density
- Insufficient resolution in regions of rapid expression change
- Loss of stochastic modeling of temporal uncertainty

**6. Connection to Differential Equation Framework**

**Bridging Observation and Inference:**

Stick-breaking provides the essential bridge for inverting the observation process:

\begin{align}
\text{Differential Equation:} \quad &t^* \rightarrow \text{Gene Expression} \rightarrow \text{Counts} \\
\text{Inference Problem:} \quad &\text{Counts} \rightarrow \text{Gene Expression} \rightarrow t^*
\end{align}

The stick-breaking construction enables tractable Bayesian inference over the biologically meaningful temporal manifold by:

**Preserving Dynamical Structure:**

The ordered constraint ensures that inferred temporal coordinates respect the underlying dynamical system structure, maintaining consistency with the forward differential equation model.

**Enabling Posterior Inference:**

The explicit probability density over $\mathcal{T}_N^{T_{\max}}$ enables efficient MCMC sampling and variational inference for temporal coordinate estimation.

**Maintaining Analytical Tractability:**

The construction preserves the analytical advantages of the piecewise solutions while providing a principled probabilistic framework for temporal uncertainty quantification.

**Validation Connection:**

The effectiveness of this approach is demonstrated in the parameter recovery validation studies (see §\ref{sec-validation-strategy}), where stick-breaking temporal coordinates enable accurate recovery of both kinetic and temporal parameters across diverse biological scenarios.
